---
title: "NSDUH 2024 Data Viz Project"
author: "Kat Vrotsos & Rachel Byers"
format: 
  dashboard:
    theme: 
      light: minty
      dark: solar
embed-resources: true
orientation: columns
---

# Data {.sidebar}

These charts were created from the 2024 National Survey on Drug Use and Health (NSDUH). NSDUH began in 1971 and provides important information to guide and support prevention and treatment programs. NSDUH provides data for monitoring substance use trends and provides critical information to inform public health policy. Learn more about NSDUH at <https://nsduhweb.rti.org/respweb/homepage.cfm>

These figures were created with unweighted survey data. In order to interpret the values as nationally representative, complex survey design, including survey weights, must be applied to the data. Our figures are only intended to display our ability to visualize data for the purposes of this class.

```{r}
#| label: install-packages
#| echo: false
#install.packages('tidyverse')
#install.packages('here')
#install.packages('rio')
#install.packages('janitor')
#install.packages('survey')
#install.packages('ggplot2')
```

```{r}
#| label: load-packages
#| echo: false
library(tidyverse)
library(here)
library(rio)
library(janitor)
library(survey)
library(ggplot2)
library(dplyr)
library(RColorBrewer)
library(ggrepel)
```

```{r}
#| label: load-data
#| echo: false

nsduh2024 <- import(here("NSDUH_2024.RData"), trust = TRUE) |>
  as_tibble() |>
  clean_names() |>
    select(questid2, analwt2_c, vestr_c, verep, catag6, irsex, newrace2, eduhighcat, govtprog, income, poverty3, health2, iralcage, irmjage, irstmnmage, illemflag2, stmnmlif, irpyud5alc, irpyud5mrj, ud5cnsany, ud5opiany2, sutnedalcpy2, sutrtalcpy2, sutrtosvpy2)

```

```{r}
#| label: format-data
#| echo: false

# add labels, values, and turn into factors or character variables as needed

# dataset with count for aud and cud and age groups as a factor in nsduh2024 and nsduh_sud

nsduh2024$catag6 <- factor(
  nsduh2024$catag6,
  levels = c('1', '2', '3', '4', '5', '6'),
  labels = c(
    '12-17 Years Old', 
    '18-25 Years Old', 
    '26-34 Years Old', 
    '34-49 Years Old',
    '50-64 Years Old',
    '65 or Older')
)

nsduh_sud <- nsduh2024 |>
  select(catag6, irpyud5alc, irpyud5mrj) |>
  mutate(
    aud_count = sum(irpyud5alc),
    cud_count = sum(irpyud5mrj)    
  )

nsduh_sud_agegroup <- nsduh2024 |>
  group_by(catag6) |>
  summarize(
    aud_count = sum(irpyud5alc),
    cud_count = sum(irpyud5mrj)    
  )

# past-year AUD as a factor

nsduh2024$irpyud5alc <- factor(
  nsduh2024$irpyud5alc,
  levels = c("0", "1"),
  labels = c("No", "Yes")
)

# poverty level as a factor

nsduh2024$poverty3 <- factor(
  nsduh2024$poverty3,
  levels = c(".", "1", "2", "3"),           # Order of levels
  labels = c("College Dorm", "Below Poverty Threshold", "Income up to 2x Threshold", "Income up to 3x Threshold") # Labels for each level
)

# race/ethnicity as a factor

nsduh2024$newrace2 <- factor(
  nsduh2024$newrace2,
  levels = c("1", "2", "3", "4", "5", "6", "7"),           # Order of levels
  labels = c("White", "Black/African American", "Native American or Alaskan Native", "Native Hawaiian or Other Pacific Island", "Asian", "Multiracial", "Hispanic") # Labels for each level
)

# past-year CUD as a factor

nsduh2024$irpyud5mrj <- factor(
  nsduh2024$irpyud5mrj,
  levels = c('0', '1'),
  labels = c('No', 'Yes')
)

```

# Sample by Race/Ethnicity

## Column (.panel width="70%")

### Final Version

```{r}
#| title: 'The complete 2024 NSDUH data set is very large and provides a wealth of information. For the purposes of this assignment, we are using only a portion of the data. The visual below provides a breakdown of race/ethnicity of participants in our portion of the data.'
#| fig-width: 10
#| fig-height: 7
#| out-width: '100%'
  
finaldemo <- nsduh2024 %>%
  drop_na(newrace2) %>%
  count(newrace2) %>%
  mutate(newrace2 = fct_reorder(newrace2, n)) %>%
  ggplot(aes(x = n, y = newrace2)) +
  geom_col(fill = "#1B9E77") +
  labs(title = "Sample Race/Ethnicity", x = "Race/Ethnicity", y = "Count", caption = 'Source: 2024 NSDUH') +
  theme_minimal()+
  theme(panel.grid.major.x = element_blank(),
        plot.title = element_text(size = 16, face = "bold"),
        axis.title = element_text(size = 12, face = "bold"),
        axis.text.x = element_text(size = 10),
        plot.caption = element_text(size = 9, color = "gray50", hjust = 0))
finaldemo
```

## Column {.tabset width="30%"}

### Draft 1

```{r}
#| fig-width: 10
#| fig-height: 7
#| out-width: '100%'

ggplot(nsduh2024, aes(x = newrace2)) +
  geom_bar(fill = "steelblue") +
  theme_minimal() +
  labs(title = "Count of Race/Ethnicity", x = "Race/Ethnicity", y = "Count")
```

### Draft 2

```{r}
#| fig-width: 10
#| fig-height: 7
#| out-width: '100%'

#edit new visual, I had to put this as a placeholder for the tab name to work properly

ggplot(nsduh2024, aes(x = newrace2)) +
  geom_bar(fill = "steelblue") +
  theme_minimal() +
  labs(title = "2024 NSDUH Data Subset Race/Ethnicity", x = "Race/Ethnicity", y = "Count") +
  theme(axis.text.x = element_text(
      angle = 45,        
      hjust = 1,        
      vjust = 1))

```

# Alocohol Use Disorder and Poverty Status

## Column (.panel width="70%")

### Final Version

```{r}
#| title: 'One of the demographic pieces of information captured by the NSDUH is poverty status. The data set also provides information on alcohol use disorder. We wanted to see the breakdown of alcohol use disorder by poverty status, which is shown below.'
#| fig-width: 10
#| fig-height: 7
#| out-width: '100%'

ggplot(na.omit(nsduh2024), aes(x = poverty3, fill = irpyud5alc)) +
  geom_bar() +
  theme_minimal() +
  labs(title = "Alcohol Use Disorder by Poverty Level", x = "Poverty Level", y = "Count",
       fill = "Alcohol Use Disorder") +
  scale_fill_brewer(palette = "Dark2") +
  theme(legend.position = "none",
        panel.grid.major.x = element_blank(),
        plot.title = element_text(size = 16, face = "bold"),
        axis.title = element_text(size = 12, face = "bold"),
        axis.text.x = element_text(size = 10),
        plot.caption = element_text(size = 9, color = "gray50", hjust = 0))

#use to add percentage labels??


#nsduh2024AUlabels = nsduh2024 %>%
  #arrange(irpyud5alc, desc(value)) %>%
  #group_by(irpyud5alc) %>%
  #mutate(label_y = 0.5 * cumsum(pct),
         #label_y = cumsum(label_y)) %>%
  #ungroup()

```

## Column {.tabset width="30%"}

### Draft 1

```{r}
#| title: 'Alcohol Use Disorder by Poverty level'
#| fig-width: 10
#| fig-height: 7
#| out-width: '100%'

ggplot(na.omit(nsduh2024), aes(x = poverty3, fill = irpyud5alc)) +
  geom_bar() +
  theme_minimal() +
  labs(title = "Alcohol Use Disorder by Poverty Level", x = "Poverty Level", y = "Count",
       fill = "Alcohol Use Disorder")
```

### Draft 2

```{r}
#| fig-width: 10
#| fig-height: 7
#| out-width: '100%'

ggplot(na.omit(nsduh2024), aes(x = poverty3, fill = irpyud5alc)) +
  geom_bar() +
  theme_minimal() +
  labs(title = "Alcohol Use Disorder by Poverty Level", x = "Poverty Level", y = "Count",
       fill = "Alcohol Use Disorder") +
  scale_fill_brewer(palette = "Dark2") +
  theme(legend.position = "none",
        panel.grid.major.x = element_blank(),
        plot.title = element_text(size = 16, face = "bold"),
        axis.title = element_text(size = 12, face = "bold"),
        axis.text.x = element_text(size = 10),
        plot.caption = element_text(size = 9, color = "gray50", hjust = 0))

```

# Alcohol Use Disorder by Age Group

```{r}
#kat
#setting up

# get proportion of age group with aud
aud_pct_age <- nsduh2024 |>
  group_by(catag6, irpyud5alc) |>
  summarize(aud_count = n()) |>
  mutate(aud_pct = aud_count / sum(aud_count)) |>
  filter(irpyud5alc == "Yes")

aud_pct_age_f <- aud_pct_age

# shorten catag6 values for final viz
aud_pct_age_f$catag6 <- recode(
  aud_pct_age_f$catag6,
    '12-17 Years Old' = '12 to 17', 
    '18-25 Years Old' = '18 to 25', 
    '26-34 Years Old' = '26 to 34', 
    '34-49 Years Old' = '34 to 49',
    '50-64 Years Old' = '50 to 64',
    '65 or Older' = '65+')

```

## Column (.panel width="70%")

### Final Version

```{r}
#| title: Past-Year Alcohol Use Disorder by Age Group in 2024
#| fig-width: 10
#| fig-height: 7
#| out-width: '100%'

aud_pct_age_f |>
  ggplot(aes(catag6, aud_pct, fill = catag6)) +
  geom_col(position = 'dodge') +
  geom_label_repel(aes(label = scales::percent(aud_pct, accuracy = 0.1)),
                   nudge_y = 0.01,
                   fill = NA) +
  scale_y_continuous(labels = scales::percent) +
  #coord_flip() +
  labs(title = 'Past-Year Alcohol Use Disorder by Age Group in 2024',
       caption = 'Source: 2024 NSDUH',
       x = 'Age (Years)',
       y = 'Past Year AUD (Percentage)') +
  theme_minimal() +
  scale_fill_brewer(palette = "Dark2") +
  theme(legend.position = "none",
        panel.grid.major.x = element_blank(),
        plot.title = element_text(size = 16, face = "bold"),
        axis.title = element_text(size = 12, face = "bold"),
        axis.text.x = element_text(size = 10),
        #axis.text.y = element_text(size = 10),
        plot.caption = element_text(size = 9, color = "gray50", hjust = 0)
       )
```

## Column {.tabset width="30%"}

### About

Description of how these were scored from codebook based on DSM-5

### Draft 1

```{r}
#| title: Number of People with Past-Year Alcohol Use Disorder by Age Group in 2024
#| fig-width: 10
#| fig-height: 7
#| out-width: '100%'

nsduh_sud |>
  group_by(catag6) |>
  ggplot(aes(catag6, aud_count)) +
  geom_col() +
  labs(title = 'Number of People with Past-Year Alcohol Use Disorder by Age Group in 2024',
       x = 'Age Group',
       y = 'Count') +
  theme_minimal()

```

### Draft 2

```{r}
#| title: Proportion of People with Past-Year Alcohol Use Disorder in 2024 by Age Group
#| fig-width: 10
#| fig-height: 7
#| out-width: '100%'

aud_pct_age |>
  ggplot(aes(catag6, aud_pct)) +
  geom_col(position = 'dodge') +
  scale_y_continuous(labels = scales::percent) +
  coord_flip() +
  labs(title = 'Proportion of People with Past-Year Alcohol Use Disorder in 2024 by Age Group',
       x = 'Age Group',
       y = 'Percentage with AUD') +
  theme_minimal()

```

# Received Alcohol Use Treatment if Needed

```{r}
# kat
# setting up

# turn classified as needing substance use treatment for alc use - past year into a factor

nsduh2024$sutnedalcpy2 <- factor(
  nsduh2024$sutnedalcpy2,
  levels = c('0', '1'),
  labels = c('No', 'Yes'))

# turn received substance use treatment for alc use - past year into a factor

nsduh2024$sutrtalcpy2 <- factor(
  nsduh2024$sutrtalcpy2,
  levels = c('0', '1'),
  labels = c('No', 'Yes'))

#dataset with people who needed su treatment for alc use in the past year
nsduh_need_alc <- nsduh2024 |>
  filter(sutnedalcpy2 == 'Yes') |>
  group_by(sutrtalcpy2) |>
  summarize(treat_count = n()) |>
  mutate(treat_pct = treat_count / sum(treat_count))

```

## Column (.panel width="70%")

### Final Version

```{r}
#| title: 'Past-Year Alcohol Use Treatment Needs Met in 2024'
#| fig-width: 10
#| fig-height: 7
#| out-width: '100%'

nsduh_need_alc |>
  ggplot(aes(sutrtalcpy2, treat_pct, fill = sutrtalcpy2)) +
  geom_col() +
  scale_y_continuous(labels = scales::percent) +
  geom_label_repel(aes(label = scales::percent(treat_pct, accuracy = 0.1)),
                   nudge_y = 0.05,
                   fill = NA) +
  labs(title = 'Past-Year Alcohol Use Treatment Needs Met in 2024',
       x = 'Received Treatment',
       y = 'Needed Treatment (Percentage)',
       caption = 'Source: 2024 NSDUH') +
  theme_minimal() +
  theme(legend.position = "none",
        panel.grid.major.x = element_blank(),
        plot.title = element_text(size = 16, face = "bold"),
        axis.title = element_text(size = 12, face = "bold"),
        axis.text.x = element_text(size = 10),
        #axis.text.y = element_text(size = 10),
        plot.caption = element_text(size = 9, color = "gray50", hjust = 0)
       )

```

## Column {.tabset width="30%"}

### About

Description of how needing treatment is computed

### Draft 1

```{r}
#| title: 'Past-Year Alcohol Use Treatment Needs Met in 2024'
#| fig-width: 10
#| fig-height: 7
#| out-width: '100%'

nsduh_need_alc |>
  ggplot(aes(sutrtalcpy2, treat_pct)) +
  geom_col() +
  scale_y_continuous(labels = scales::percent) +
  labs(title = 'Past-Year Alcohol Use Treatment Needs Met in 2024',
       subtitle = 'Among individuals who were classified as needing treatment for their alcohol use in the 2024 NSDUH',
       x = 'Received Treatment',
       y = 'Treatment Needed (Percentage)') +
  theme_minimal()

```

### Draft 2

```{r}
#| fig-width: 10
#| fig-height: 7
#| out-width: '100%'

nsduh_need_alc |>
  ggplot(aes(sutrtalcpy2, treat_pct, fill = sutrtalcpy2)) +
  geom_col() +
  scale_y_continuous(labels = scales::percent) +
  geom_label_repel(aes(label = scales::percent(treat_pct, accuracy = 0.1)),
                   nudge_y = 0.05,
                   fill = NA) +
  labs(title = 'Past-Year Alcohol Use Treatment Needs Met in 2024',
       x = 'Received Treatment',
       y = 'Treatment Needed (Percentage)') +
  theme_minimal() +
  theme(legend.position = "none")

```
