---
title: "NSDUH 2024 Data Viz Project"
author: "Kat Vrotsos & Rachel Byers"
format: 
  dashboard:
    theme: 
      light: minty
      dark: solar
embed-resources: true
orientation: columns
---

```{r}
#| label: install-packages
#| echo: false
#install.packages('tidyverse')
#install.packages('here')
#install.packages('rio')
#install.packages('janitor')
#install.packages('survey')
#install.packages('ggplot2')
```

```{r}
#| label: load-packages
#| echo: false
library(tidyverse)
library(here)
library(rio)
library(janitor)
library(survey)
library(ggplot2)
library(dplyr)
library(RColorBrewer)
library(ggrepel)
library(scales)
```

```{r}
#| label: load-data
#| echo: false

nsduh2024 <- import(here("NSDUH_2024.RData"), trust = TRUE) |>
  as_tibble() |>
  clean_names() |>
    select(questid2, analwt2_c, vestr_c, verep, catag6, irsex, newrace2, eduhighcat, govtprog, income, poverty3, health2, iralcage, irmjage, irstmnmage, illemflag2, stmnmlif, irpyud5alc, irpyud5mrj, ud5cnsany, ud5opiany2, sutnedalcpy2, sutrtalcpy2, sutrtosvpy2)

```

```{r}
#| label: format-data
#| echo: false

# add labels, values, and turn into factors or character variables as needed

# dataset with count for aud and cud and age groups as a factor in nsduh2024 and nsduh_sud

nsduh2024$catag6 <- factor(
  nsduh2024$catag6,
  levels = c('1', '2', '3', '4', '5', '6'),
  labels = c(
    '12-17 Years Old', 
    '18-25 Years Old', 
    '26-34 Years Old', 
    '34-49 Years Old',
    '50-64 Years Old',
    '65 or Older')
)

nsduh_sud <- nsduh2024 |>
  select(catag6, irpyud5alc, irpyud5mrj) |>
  mutate(
    aud_count = sum(irpyud5alc),
    cud_count = sum(irpyud5mrj)    
  )

nsduh_sud_agegroup <- nsduh2024 |>
  group_by(catag6) |>
  summarize(
    aud_count = sum(irpyud5alc),
    cud_count = sum(irpyud5mrj)    
  )

# past-year AUD as a factor

nsduh2024$irpyud5alc <- factor(
  nsduh2024$irpyud5alc,
  levels = c("0", "1"),
  labels = c("No", "Yes")
)

# poverty level as a factor

nsduh2024$poverty3 <- factor(
  nsduh2024$poverty3,
  levels = c(".", "1", "2", "3"),           # Order of levels
  labels = c("College Dorm", "Below Poverty Threshold", "Income up to 2x Threshold", "Income up to 3x Threshold") # Labels for each level
)

# race/ethnicity as a factor

nsduh2024$newrace2 <- factor(
  nsduh2024$newrace2,
  levels = c("1", "2", "3", "4", "5", "6", "7"),           # Order of levels
  labels = c("White", "Black/African American", "Native American or Alaskan Native", "Native Hawaiian or Other Pacific Island", "Asian", "Multiracial", "Hispanic") # Labels for each level
)

```

# 2024 NSDUH Data

## Column {.tabset title="2024 NSDUH Data" width="70%"}

### About the Data

2024 National Survey on Drug Use and Health (NSDUH) data was used to create this dashboard. NSDUH began in 1971 and provides important information to guide and support prevention and treatment programs. The target population includes civilian, non-institutionalized people ages 12 or older that live in the United States (N = 58,633). NSDUH provides data for monitoring substance use trends and provides critical information to inform public health policy. Learn more about NSDUH at <https://nsduhweb.rti.org/respweb/homepage.cfm>

These figures were created with **unweighted survey data**. In order to interpret the values as nationally representative, complex survey design, including survey weights, must be applied to the data. Our figures are only intended to display our ability to visualize data for the purposes of this class. We used imputed and/or recoded variables as suggested in the [2024 NSDUH Public Use File Data Users’ Guide](https://www.samhsa.gov/data/sites/default/files/reports/rpt56775/2024-nsduh-puf-data-users-guide.pdf) to reduce bias related to missing and non-response data.

### Alcohol Use Disorder Classification

The NSDUH'S past-year alcohol use disorder (AUD) measure was based on the criteria from the *Diagnostic and Statistical Manual of Mental Disorders*, 5th edition. Participants were classified as having AUD if they met two or more of the 11 criteria listed below:

1.  Ended up drinking alcohol many times in larger amounts or for a longer time than
    intended

2.  One or more times tried and was unable to cut down or stop drinking alcohol or
    wished could cut down or stop drinking alcohol

3.  Spent a great deal of time drinking alcohol, feeling its effects, or getting over the
    effects of drinking, or trying to get alcohol

4.  Wanted to use alcohol so badly could not think of anything else or had a strong urge to use alcohol

5.  Serious problems at work, school, or home, such as: missing a lot of work or school; getting demoted, having your hours cut, or losing a job; not being able to get a job or keep a job; getting suspended, expelled, or dropping out of school; and failing to take care of family

6.  Problems with family or friends that were caused or made worse by using alcohol and continued to use alcohol even though it often caused arguments or problems with family or friends

7.  Gave up or spent a lot less time doing important social, occupational, or recreational
    activities because of alcohol use

8.  Repeatedly got into situations where drinking alcohol increased your chances of
    getting physically hurt

9.  Continued to use alcohol even though it was causing problems with emotions, mental
    health, or physical health, or drinking gave you repeated blackouts

10. Needed to drink a lot more alcohol to get the desired feeling, or drinking the same
    amount had less of an effect

11. Reported experiencing two or more alcohol withdrawal symptoms after alcohol use
    was cut back or stopped or respondent used the same or similar substance to avoid
    having alcohol withdrawal symptoms during the past 12 months. Symptoms include: (i) sweating or feeling that heart was beating fast; (ii) having hands tremble; (iii) having trouble sleeping; (iv) vomiting or feeling nauseous; (v) seeing, hearing, or feeling things that were not really there; (vi) feeling like could not sit still; (vii) feeling anxious; and (viii) having seizures or fits. Substances used to avoid withdrawal symptoms could include alcohol or prescription sedatives, tranquilizers, sleeping pills, or downers.

If participants reported alcohol use in the past 12 months and either used alcohol on six or more days in the past year or did not reply to the item, then they were asked questions about past-year AUD. Additional information on the AUD measure is available in the [2024 NSDUH Public Use File Codebook](https://www.samhsa.gov/data/system/files/media-puf-file/nsduh-2024-ds0001-info-codebook_v1.pdf).

## Column {.tabset title="Variables" width="30%"}

### Race/Ethnicity

### AUD

Count Percentage

### Poverty Status

### Age

# Sample by Race/Ethnicity

## Column (.panel width="70%")

### Final Version

```{r}
#| title: 'The visual below provides a breakdown of race/ethnicity of participants from the 2024 NSDUH.'
#| fig-width: 10
#| fig-height: 7
#| out-width: '100%'
  
finaldemo <- nsduh2024 %>%
  drop_na(newrace2) %>%
  count(newrace2) %>%
  mutate(newrace2 = fct_reorder(newrace2, n)) %>%
  ggplot(aes(x = n, y = newrace2)) +
  geom_col(fill = "#1B9E77") +
  labs(title = "Sample Race/Ethnicity", x = "Count", y = "Race/Ethnicity", caption = 'Source: 2024 NSDUH') +
  theme_minimal()+
  theme(panel.grid.major.x = element_blank(),
        plot.title = element_text(size = 16, face = "bold"),
        axis.title = element_text(size = 14, face = "bold"),
        axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12),
        plot.caption = element_text(size = 9, color = "gray50", hjust = 0))

finaldemo
```

## Column {.tabset width="30%"}

### Draft 1

```{r}
#| fig-width: 10
#| fig-height: 7
#| out-width: '100%'

ggplot(nsduh2024, aes(x = newrace2)) +
  geom_bar(fill = "steelblue") +
  theme_minimal() +
  labs(title = "Count of Race/Ethnicity", x = "Race/Ethnicity", y = "Count")
```

### Draft 2

```{r}
#| fig-width: 10
#| fig-height: 7
#| out-width: '100%'

#edit new visual, I had to put this as a placeholder for the tab name to work properly

ggplot(nsduh2024, aes(x = newrace2)) +
  geom_bar(fill = "steelblue") +
  theme_minimal() +
  labs(title = "2024 NSDUH Data Subset Race/Ethnicity", x = "Race/Ethnicity", y = "Count") +
  theme(axis.text.x = element_text(
      angle = 45,        
      hjust = 1,        
      vjust = 1))

```

# AUD and Poverty Status

## Column (.panel width="70%")

### Final Version

```{r}
#| title: 'One of the demographic pieces of information captured by the NSDUH is poverty status. The data set also provides information on alcohol use disorder. We wanted to see the breakdown of alcohol use disorder by poverty status, which is shown below.'
#| fig-width: 10
#| fig-height: 7
#| out-width: '100%'

ggplot(na.omit(nsduh2024), aes(x = poverty3, fill = irpyud5alc)) +
  geom_bar() +
  theme_minimal() +
  labs(title = "Alcohol Use Disorder by Poverty Level", x = "Poverty Level", y = "Count",
       fill = "Alcohol Use Disorder") +
  scale_fill_brewer(palette = "Dark2") +
  theme(legend.position = "none",
        panel.grid.major.x = element_blank(),
        plot.title = element_text(size = 16, face = "bold"),
        axis.title = element_text(size = 14, face = "bold"),
        axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12),
        plot.caption = element_text(size = 9, color = "gray50", hjust = 0))

#use to add percentage labels??


#nsduh2024AUlabels = nsduh2024 %>%
  #arrange(irpyud5alc, desc(value)) %>%
  #group_by(irpyud5alc) %>%
  #mutate(label_y = 0.5 * cumsum(pct),
         #label_y = cumsum(label_y)) %>%
  #ungroup()

```

## Column {.tabset width="30%"}

### Draft 1

```{r}
#| fig-width: 10
#| fig-height: 7
#| out-width: '100%'

ggplot(na.omit(nsduh2024), aes(x = poverty3, fill = irpyud5alc)) +
  geom_bar() +
  theme_minimal() +
  labs(title = "Alcohol Use Disorder by Poverty Level", x = "Poverty Level", y = "Count",
       fill = "Alcohol Use Disorder")
```

### Draft 2

```{r}
#| fig-width: 10
#| fig-height: 7
#| out-width: '100%'

ggplot(na.omit(nsduh2024), aes(x = poverty3, fill = irpyud5alc)) +
  geom_bar() +
  theme_minimal() +
  labs(title = "Alcohol Use Disorder by Poverty Level", x = "Poverty Level", y = "Count",
       fill = "Alcohol Use Disorder") +
  scale_fill_brewer(palette = "Dark2") +
  theme(legend.position = "none",
        panel.grid.major.x = element_blank(),
        plot.title = element_text(size = 16, face = "bold"),
        axis.title = element_text(size = 12, face = "bold"),
        axis.text.x = element_text(size = 10),
        plot.caption = element_text(size = 9, color = "gray50", hjust = 0))

```

# AUD by Age Group

```{r}
#kat
#setting up

# get proportion of age group with aud
aud_pct_age <- nsduh2024 |>
  group_by(catag6, irpyud5alc) |>
  summarize(aud_count = n()) |>
  mutate(aud_pct = aud_count / sum(aud_count)) |>
  filter(irpyud5alc == "Yes")

aud_pct_age_f <- aud_pct_age

# shorten catag6 values for final viz
aud_pct_age_f$catag6 <- recode(
  aud_pct_age_f$catag6,
    '12-17 Years Old' = '12 to 17', 
    '18-25 Years Old' = '18 to 25', 
    '26-34 Years Old' = '26 to 34', 
    '34-49 Years Old' = '34 to 49',
    '50-64 Years Old' = '50 to 64',
    '65 or Older' = '65+')

```

## Column (.panel width="70%")

### Final Version

```{r}
#| title: Past-Year Alcohol Use Disorder by Age Group in 2024
#| fig-width: 10
#| fig-height: 7
#| out-width: '100%'

aud_pct_age_f |>
  ggplot(aes(catag6, aud_pct, fill = catag6)) +
  geom_col(position = 'dodge') +
  geom_label_repel(aes(label = scales::percent(aud_pct, accuracy = 0.1)),
                   nudge_y = 0.01,
                   fill = NA) +
  scale_y_continuous(labels = scales::percent) +
  #coord_flip() +
  labs(title = 'Past-Year Alcohol Use Disorder by Age Group in 2024',
       caption = 'Source: 2024 NSDUH',
       x = 'Age (Years)',
       y = 'Past Year AUD (Percentage)') +
  theme_minimal() +
  scale_fill_brewer(palette = "Dark2") +
  theme(legend.position = "none",
        panel.grid.major.x = element_blank(),
        plot.title = element_text(size = 16, face = "bold"),
        axis.title = element_text(size = 12, face = "bold"),
        axis.text.x = element_text(size = 10, face = "bold"),
        #axis.text.y = element_text(size = 10),
        plot.caption = element_text(size = 9, color = "gray50", hjust = 0)
       )
```

## Column {.tabset width="30%"}

### About

This figure shows the percentage of participants that were classified as having alcohol use disorder in the past year within each age group.

### Draft 1

```{r}
#| title: Number of People with Past-Year Alcohol Use Disorder by Age Group in 2024
#| fig-width: 10
#| fig-height: 7
#| out-width: '100%'

nsduh_sud |>
  group_by(catag6) |>
  ggplot(aes(catag6, aud_count)) +
  geom_col() +
  labs(title = 'Number of People with Past-Year Alcohol Use Disorder by Age Group in 2024',
       x = 'Age Group',
       y = 'Count') +
  theme_minimal()

```

### Draft 2

```{r}
#| title: Proportion of People with Past-Year Alcohol Use Disorder in 2024 by Age Group
#| fig-width: 10
#| fig-height: 7
#| out-width: '100%'

aud_pct_age |>
  ggplot(aes(catag6, aud_pct)) +
  geom_col(position = 'dodge') +
  scale_y_continuous(labels = scales::percent) +
  coord_flip() +
  labs(title = 'Proportion of People with Past-Year Alcohol Use Disorder in 2024 by Age Group',
       x = 'Age Group',
       y = 'Percentage with AUD') +
  theme_minimal()

```

# Alcohol Use Treatment Needs

```{r}
# kat
# setting up

# turn classified as needing substance use treatment for alc use - past year into a factor

nsduh2024$sutnedalcpy2 <- factor(
  nsduh2024$sutnedalcpy2,
  levels = c('0', '1'),
  labels = c('No', 'Yes'))

# turn received substance use treatment for alc use - past year into a factor

nsduh2024$sutrtalcpy2 <- factor(
  nsduh2024$sutrtalcpy2,
  levels = c('0', '1'),
  labels = c('No', 'Yes'))

#dataset with people who needed su treatment for alc use in the past year
nsduh_need_alc <- nsduh2024 |>
  filter(sutnedalcpy2 == 'Yes') |>
  group_by(sutrtalcpy2) |>
  summarize(treat_count = n()) |>
  mutate(treat_pct = treat_count / sum(treat_count))

```

## Column (.panel width="70%")

### Final Version

```{r}
#| title: 'Past-Year Alcohol Use Treatment Needs Met in 2024'
#| fig-width: 10
#| fig-height: 7
#| out-width: '100%'

nsduh_need_alc |>
  ggplot(aes(sutrtalcpy2, treat_pct, fill = sutrtalcpy2)) +
  geom_col() +
  scale_y_continuous(labels = scales::percent) +
  geom_label_repel(aes(label = scales::percent(treat_pct, accuracy = 0.1)),
                   nudge_y = 0.05,
                   fill = NA) +
  labs(title = 'Past-Year Alcohol Use Treatment Needs Met in 2024',
       x = 'Received Treatment',
       y = 'Needed Treatment (Percentage)',
       caption = 'Source: 2024 NSDUH') +
  theme_minimal() +
  scale_fill_brewer(palette = "Paired") +
  theme(legend.position = "none",
        panel.grid.major.x = element_blank(),
        plot.title = element_text(size = 16, face = "bold"),
        axis.title = element_text(size = 12, face = "bold"),
        axis.text.x = element_text(size = 12, face = "bold", color = "black"),
        #axis.text.y = element_text(size = 10),
        plot.caption = element_text(size = 9, color = "gray50", hjust = 0)
       )

```

## Column {.tabset width="30%"}

### About

This figure shows the percentage of people who received substance use treatment for their alcohol use among participants who were classified as needing treatment (i.e., having alcohol use disorder).

### Draft 1

```{r}
#| title: 'Past-Year Alcohol Use Treatment Needs Met in 2024'
#| fig-width: 10
#| fig-height: 7
#| out-width: '100%'

nsduh_need_alc |>
  ggplot(aes(sutrtalcpy2, treat_pct)) +
  geom_col() +
  scale_y_continuous(labels = scales::percent) +
  labs(title = 'Past-Year Alcohol Use Treatment Needs Met in 2024',
       subtitle = 'Among individuals who were classified as needing treatment for their alcohol use in the 2024 NSDUH',
       x = 'Received Treatment',
       y = 'Treatment Needed (Percentage)') +
  theme_minimal()

```

### Draft 2

```{r}
#| fig-width: 10
#| fig-height: 7
#| out-width: '100%'

nsduh_need_alc |>
  ggplot(aes(sutrtalcpy2, treat_pct, fill = sutrtalcpy2)) +
  geom_col() +
  scale_y_continuous(labels = scales::percent) +
  geom_label_repel(aes(label = scales::percent(treat_pct, accuracy = 0.1)),
                   nudge_y = 0.05,
                   fill = NA) +
  labs(title = 'Past-Year Alcohol Use Treatment Needs Met in 2024',
       x = 'Received Treatment',
       y = 'Treatment Needed (Percentage)') +
  theme_minimal() +
  theme(legend.position = "none")

```

# Our Process

## Column (.panel)

**Major Challenges**

-   Complex survey design (i.e., sample weights and variance estimation)

-   .sav data with formatting was too big (\> 100MB) for GitHub

**Major Victories**

-   Figured out how to format visualizations so they're not too small to read on the dashboard

-   Used the .RData file and formatted variables ourselves
