---
title: "NSDUH 2024 Data Viz Project"
author: "Kat Vrotsos & Rachel Byers"
format: 
  dashboard:
    theme: 
      light: minty
      dark: solar
embed-resources: true
orientation: rows
---

#  {.sidebar}

These charts are created from the 2024 National Survey on Drug Use and Health (NSDUH). NSDUH began in 1971 and provides important information to guide and support prevention and treatment programs. NSDUH provides data for monitoring substance use trends and provides critical information to inform public health policy. Learn more about NSDUH at https://nsduhweb.rti.org/respweb/homepage.cfm

These figures were created with unweighted survey data. In order to interpret the values as nationally representative, complex survey design, including survey weights, must be applied to the data. Our figures are only intended to display our ability to visualize data for the purposes of this class.

```{r}
#| label: install-packages
#install.packages('tidyverse')
#install.packages('here')
#install.packages('rio')
#install.packages('janitor')
#install.packages('survey')
#install.packages('ggplot2')
```

```{r}
#| label: load-packages
library(tidyverse)
library(here)
library(rio)
library(janitor)
library(survey)
library(ggplot2)
library(dplyr)
```

```{r}
#| label: load-data

nsduh2024 <- import(here("NSDUH_2024.RData"), trust = TRUE) |>
  as_tibble() |>
  clean_names() |>
    select(questid2, analwt2_c, vestr_c, verep, catag6, irsex, newrace2, eduhighcat, govtprog,
         income, poverty3, health2, iralcage, irmjage, irstmnmage, illemflag2, stmnmlif,
         irpyud5alc, irpyud5mrj, ud5cnsany, ud5opiany2)

```

```{r}
#| label: format-data
#| echo: false

# add labels, values, and turn into factors or character variables as needed

# dataset with count for aud and cud and age groups as a factor in nsduh2024 and nsduh_sud

nsduh2024$catag6 <- factor(
  nsduh2024$catag6,
  levels = c('1', '2', '3', '4', '5', '6'),
  labels = c(
    '12-17 Years Old', 
    '18-25 Years Old', 
    '26-34 Years Old', 
    '34-49 Years Old',
    '50-64 Years Old',
    '65 or Older')
)

nsduh_sud <- nsduh2024 |>
  select(catag6, irpyud5alc, irpyud5mrj) |>
  mutate(
    aud_count = sum(irpyud5alc),
    cud_count = sum(irpyud5mrj)    
  )

nsduh_sud_agegroup <- nsduh2024 |>
  group_by(catag6) |>
  summarize(
    aud_count = sum(irpyud5alc),
    cud_count = sum(irpyud5mrj)    
  )

# past-year AUD as a factor

nsduh2024$irpyud5alc <- factor(
  nsduh2024$irpyud5alc,
  levels = c("0", "1"),
  labels = c("No", "Yes")
)

# poverty level as a factor

nsduh2024$poverty3 <- factor(
  nsduh2024$poverty3,
  levels = c(".", "1", "2", "3"),           # Order of levels
  labels = c("College Dorm", "Below Poverty Threshold", "Income up to 2x Threshold", "Income up to 3x Threshold") # Labels for each level
)

# race/ethnicity as a factor

class("newrace2")
nsduh2024$newrace2 <- factor(
  nsduh2024$newrace2,
  levels = c("1", "2", "3", "4", "5", "6", "7"),           # Order of levels
  labels = c("White", "Black/African American", "Native American or Alaskan Native", "Native Hawaiian or Other Pacific Island", "Asian", "Multiracial", "Hispanic") # Labels for each level
)

# past-year CUD as a factor

nsduh2024$irpyud5mrj <- factor(
  nsduh2024$irpyud5mrj,
  levels = c('0', '1'),
  labels = c('No', 'Yes')
)

```

# Sample by Race/Ethnicity

## Row (.panel height="20%")

### About
The complete 2024 NSDUH data set is very large and provides a wealth of information. For the purposes of this assignment, we are using only a portion of the data. The visual below provides a breakdown of race/ethnicity of participants in our portion of the data.

## Row {.tabset height="80%"}

### Final Version

```{r}
ggplot(nsduh2024, aes(x = newrace2)) +
  geom_bar(fill = "steelblue") +
  theme_minimal() +
  labs(title = "2024 NSDUH Data Subset Race/Ethnicity", x = "Race/Ethnicity", y = "Count") +
  theme(axis.text.x = element_text(
      angle = 45,        
      hjust = 1,        
      vjust = 1))
```

### Draft 2

```{r}
#| title: 'needs to be updated'

#edit new visual, I had to put this as a placeholder for the tab name to work properly

ggplot(nsduh2024, aes(x = newrace2)) +
  geom_bar(fill = "steelblue") +
  theme_minimal() +
  labs(title = "2024 NSDUH Data Subset Race/Ethnicity", x = "Race/Ethnicity", y = "Count") +
  theme(axis.text.x = element_text(
      angle = 45,        
      hjust = 1,        
      vjust = 1))

```

### Draft 1

```{r}
#| title: 'First Visual: Breakdown of Sample by Race/Ethnicity'

ggplot(nsduh2024, aes(x = newrace2)) +
  geom_bar(fill = "steelblue") +
  theme_minimal() +
  labs(title = "Count of Race/Ethnicity", x = "Race/Ethnicity", y = "Count")
```


# Alcohol Use Disorder and Poverty

## Row (.panel height="20%")

### About
Description

## Row {.tabset height="80%"}

### Final Version

```{r}

```

### Draft 2

```{r}

```

### Draft 1

```{r}
#| title: 'Alcohol Use Disorder by Poverty level'

ggplot(na.omit(nsduh2024), aes(x = poverty3, fill = irpyud5alc)) +
  geom_bar() +
  theme_minimal() +
  labs(title = "Alcohol Use Disorder by Poverty Level", x = "Poverty Level", y = "Count",
       fill = "Alcohol Use Disorder")
```

# Prevalence of Alcohol and Cannabis Use Disorders by Age Group

## Row (.panel height="20%")

### About
Description of how these were scored from codebook based on DSM-5

## Row {.tabset height="80%"}

### Final Version

```{r}

```

### Draft 2

```{r}

```

### Draft 1

```{r}
#| title: Proportion of People with Past-Year Alcohol Use Disorder in 2024 by Age Group

aud_pct_age <- nsduh2024 |>
  group_by(catag6, irpyud5alc) |>
  summarize(aud_count = n()) |>
  mutate(aud_pct = aud_count / sum(aud_count)) |>
  filter(irpyud5alc == "Yes")
  
aud_pct_age |>
  ggplot(aes(catag6, aud_pct)) +
  geom_col(position = 'dodge') +
  scale_y_continuous(labels = scales::percent) +
  coord_flip() +
  labs(title = 'Proportion of People with Past-Year Alcohol Use Disorder in 2024 by Age Group',
       x = 'Age Group',
       y = 'Percentage with AUD') +
  theme_minimal()

```

# Kat's other visual TBD

## Row (.panel height="20%")

### About
Description

## Row {.tabset height="80%"}

### Final Version

```{r}

```

### Draft 2

```{r}

```

### Draft 1

```{r}

```

