---
title: "Data Visualization Project"
author: "Kat Vrotsos and Rachel Byers"
format: dashboard
embed-resources: true
orientation: rows
---

#  {.sidebar}

These charts are created from the 2024 National Survey on Drug Use and Health (NSDUH). NSDUH began in 1971 and provides important information to guide and support prevention and treatment programs. NSDUH provides data for monitoring substance use trends and provides critical information to inform public health policy. Learn more about NSDUH at https://nsduhweb.rti.org/respweb/homepage.cfm

```{r}
#| label: install-packages
#install.packages('tidyverse')
#install.packages('here')
#install.packages('rio')
#install.packages('janitor')
#install.packages('survey')
#install.packages('ggplot2')
```

```{r}
#| label: load-packages
library(tidyverse)
library(here)
library(rio)
library(janitor)
library(survey)
library(ggplot2)
```

```{r}
#| label: load-data

nsduh2024 <- import(here("NSDUH_2024.RData"), trust = TRUE) |>
  as_tibble() |>
  clean_names() |>
    select(questid2, analwt2_c, vestr_c, verep, catag6, irsex, newrace2, eduhighcat, govtprog,
         income, poverty3, health2, iralcage, irmjage, irstmnmage, illemflag2, stmnmlif,
         irpyud5alc, irpyud5mrj, ud5cnsany, ud5opiany2)

```

```{r}
#| label: format-data

# add labels, values, and turn into factors or character variables as needed

nsduh2024 <- nsduh2024 %>%
  mutate(poverty3 = as.factor(poverty3)) %>%
  mutate(irpyud5alc = as.factor(irpyud5alc))

# conversion to factor not working?? what am I missing here??

```

# Lab PS3: Visual from Project Data

## Row {height="80%"}

```{r}
#| title: 'First Visual: Breakdown of Sample by Race/Ethnicity'

class("newrace2")
nsduh2024$newrace2 <- factor(
  nsduh2024$newrace2,
  levels = c("1", "2", "3", "4", "5", "6", "7"),           # Order of levels
  labels = c("White", "Black/African American", "Native American or Alaskan Native", "Native Hawaiian or Other Pacific Island", "Asian", "Multiracial", "Hispanic") # Labels for each level
)
ggplot(nsduh2024, aes(x = newrace2)) +
  geom_bar(fill = "steelblue") +
  theme_minimal() +
  labs(title = "Count of Race/Ethnicity", x = "Race/Ethnicity", y = "Count")
```
