---
title: "NSDUH 2024 Data Viz Project"
author: "Kat Vrotsos & Rachel Byers"
format: 
  dashboard:
    theme: 
      light: minty
      dark: solar
embed-resources: true
orientation: columns
---

```{r}
#| label: install-packages
#| echo: false
#install.packages('tidyverse')
#install.packages('here')
#install.packages('rio')
#install.packages('janitor')
#install.packages('survey')
#install.packages('ggplot2')
```

```{r}
#| label: load-packages
#| echo: false
library(tidyverse)
library(here)
library(rio)
library(janitor)
library(survey)
library(ggplot2)
library(dplyr)
library(RColorBrewer)
library(ggrepel)
library(scales)
library(ggtext)
```

```{r}
#| label: load-data
#| echo: false

nsduh2024 <- import(here("NSDUH_2024.RData"), trust = TRUE) |>
  as_tibble() |>
  clean_names() |>
    select(questid2, analwt2_c, vestr_c, verep, catag6, irsex, newrace2, eduhighcat, govtprog, income, poverty3, health2, iralcage, irmjage, irstmnmage, illemflag2, stmnmlif, irpyud5alc, irpyud5mrj, ud5cnsany, ud5opiany2, sutnedalcpy2, sutrtalcpy2, sutrtosvpy2)

```

```{r}
#| label: format-data
#| echo: false

# add labels, values, and turn into factors or character variables as needed

# dataset with count for aud and cud and age groups as a factor in nsduh2024 and nsduh_sud

nsduh2024$catag6 <- factor(
  nsduh2024$catag6,
  levels = c('1', '2', '3', '4', '5', '6'),
  labels = c(
    '12-17 Years Old', 
    '18-25 Years Old', 
    '26-34 Years Old', 
    '34-49 Years Old',
    '50-64 Years Old',
    '65 or Older')
)

nsduh_sud <- nsduh2024 |>
  select(catag6, irpyud5alc, irpyud5mrj) |>
  mutate(
    aud_count = sum(irpyud5alc),
    cud_count = sum(irpyud5mrj)    
  )

nsduh_sud_agegroup <- nsduh2024 |>
  group_by(catag6) |>
  summarize(
    aud_count = sum(irpyud5alc),
    cud_count = sum(irpyud5mrj)    
  )

# past-year AUD as a factor

nsduh2024$irpyud5alc <- factor(
  nsduh2024$irpyud5alc,
  levels = c("0", "1"),
  labels = c("No", "Yes")
)

# poverty level as a factor

nsduh2024$poverty3 <- factor(
  nsduh2024$poverty3,
  levels = c(".", "1", "2", "3"),           # Order of levels
  labels = c("College Dorm", "Below Poverty Threshold", "Income up to 2x Threshold", "Income up to 3x Threshold") # Labels for each level
)

# race/ethnicity as a factor

nsduh2024$newrace2 <- factor(
  nsduh2024$newrace2,
  levels = c("1", "2", "3", "4", "5", "6", "7"),           # Order of levels
  labels = c("White", "Black/African American", "Native American or Alaskan Native", "Native Hawaiian or Other Pacific Island", "Asian", "Multiracial", "Hispanic") # Labels for each level
)

```

# Data

## Column {.tabset title="2024 NSDUH Data"}

### About

2024 National Survey on Drug Use and Health (NSDUH) data was used to create this dashboard. NSDUH began in 1971 and provides important information to guide and support prevention and treatment programs. The target population includes civilian, non-institutionalized people ages 12 or older that live in the United States (N = 58,633). NSDUH provides data for monitoring substance use trends and provides critical information to inform public health policy. Learn more about NSDUH at <https://nsduhweb.rti.org/respweb/homepage.cfm>

These figures were created with **unweighted survey data**. In order to interpret the values as nationally representative, complex survey design, including survey weights, must be applied to the data. Our figures are only intended to display our ability to visualize data for the purposes of this class.

### Variables

We used imputed and/or recoded variables as suggested in the [2024 NSDUH Public Use File Data Users’ Guide](https://www.samhsa.gov/data/sites/default/files/reports/rpt56775/2024-nsduh-puf-data-users-guide.pdf) to reduce bias related to missing and non-response data.

| Variable | NSDUH Code | Factor Levels |
|-------------------|-------------------|----------------------------------|
| Race/Ethnicity | NEWRACE2 | White, Black/African American, Native American or Alaskan Native, Native Hawaiian or Other Pacific Island, Asian, Multiracial, or Hispanic |
| Past-Year Alcohol Use Disorder | IRPYUD5ALC | No or Yes |
| Poverty Level | POVERTY3 | Below Poverty Threshold, Income up to 2x Threshold, or Income up to 3x Threshold |
| Age Group | CATAG3 | 12-17 Years Old, 18-25 Years Old, 26-34 Years Old, 34-49 Years Old, 50-64 Years Old, or 65 or Older |
| Past-Year Alcohol Use Treatment | SUTNEDALCPY2 | No or Yes |

### Process

**Major Challenges**

-   Complex survey design (i.e., sample weights and variance estimation)

-   .sav data with formatting was too big (\> 100MB) for GitHub

**Major Victories**

-   Figured out how to format visualizations so they're not too small to read on the dashboard

-   Used the .RData file and formatted variables ourselves

## Column {.tabset title="Alcohol Use Disorder"}

### Classification

The NSDUH'S past-year alcohol use disorder (AUD) measure was based on the criteria from the *Diagnostic and Statistical Manual of Mental Disorders*, 5th edition. Participants were classified as having AUD if they met two or more of the 11 criteria listed below:

1.  Ended up drinking alcohol many times in larger amounts or for a longer time than intended

2.  One or more times tried and was unable to cut down or stop drinking alcohol or wished could cut down or stop drinking alcohol

3.  Spent a great deal of time drinking alcohol, feeling its effects, or getting over the effects of drinking, or trying to get alcohol

4.  Wanted to use alcohol so badly could not think of anything else or had a strong urge to use alcohol

5.  Serious problems at work, school, or home, such as: missing a lot of work or school; getting demoted, having your hours cut, or losing a job; not being able to get a job or keep a job; getting suspended, expelled, or dropping out of school; and failing to take care of family

6.  Problems with family or friends that were caused or made worse by using alcohol and continued to use alcohol even though it often caused arguments or problems with family or friends

7.  Gave up or spent a lot less time doing important social, occupational, or recreational activities because of alcohol use

8.  Repeatedly got into situations where drinking alcohol increased your chances of getting physically hurt

9.  Continued to use alcohol even though it was causing problems with emotions, mental health, or physical health, or drinking gave you repeated blackouts

10. Needed to drink a lot more alcohol to get the desired feeling, or drinking the same amount had less of an effect

11. Reported experiencing two or more alcohol withdrawal symptoms after alcohol use was cut back or stopped or respondent used the same or similar substance to avoid having alcohol withdrawal symptoms during the past 12 months. Symptoms include: (i) sweating or feeling that heart was beating fast; (ii) having hands tremble; (iii) having trouble sleeping; (iv) vomiting or feeling nauseous; (v) seeing, hearing, or feeling things that were not really there; (vi) feeling like could not sit still; (vii) feeling anxious; and (viii) having seizures or fits. Substances used to avoid withdrawal symptoms could include alcohol or prescription sedatives, tranquilizers, sleeping pills, or downers.

If participants reported alcohol use in the past 12 months and either used alcohol on six or more days in the past year or did not reply to the item, then they were asked questions about past-year AUD.

Additional information on NSDUH's AUD measure is available in the [2024 NSDUH Public Use File Codebook](https://www.samhsa.gov/data/system/files/media-puf-file/nsduh-2024-ds0001-info-codebook_v1.pdf).

### Count vs. Percentage

AUD was used as both count and percentage variables in different visualizations on this dashboard.

-   **Count**

    -   AUD was already a binary variable coded as 0 (*No*) and 1 (*Yes*), so it could readily be used as a count outcome

    -   Used in **AUD and Poverty Level** visualization

\

-   **Percentage**

    -   AUD was transformed into a percentage in the **AUD by Age Group** and **Alcohol Use Treatment** visualizations

    -   Additional details on computation are in located in the About tabs on both pages

# Sample by Race/Ethnicity

## Column (.panel width="70%")

### Final Version

```{r}
#| title: 'What is the breakdown of the 2024 NSDUH data by race/ethnicity?'
#| fig-width: 10
#| fig-height: 7
#| out-width: '100%'
  
finaldemo <- nsduh2024 %>%
  drop_na(newrace2) %>%
  count(newrace2) %>%
  mutate(newrace2 = fct_reorder(newrace2, n)) %>%
  ggplot(aes(x = n, y = newrace2)) +
  geom_col(fill='#1B9E77') +
  geom_label(aes(label = comma(n))) +
  labs(title = "2024 NSDUH Respondent Race/Ethnicity", x = "Count", y = "Race/Ethnicity", caption = 'Source: 2024 NSDUH') +
  scale_y_discrete(labels = label_wrap(width = 20)) +
  scale_x_continuous(labels = label_comma()) +
  theme_minimal() +
  theme(panel.grid.major.x = element_blank(),
        plot.title = element_text(size = 16, face = "bold"),
        axis.title = element_text(size = 14, face = "bold"),
        axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12),
        plot.caption = element_text(size = 10, color = "gray50", hjust = 0))

finaldemo
```

## Column {.tabset width="30%"}

### About

The visual to the left provides a breakdown of the race/ethnicity of participants from the NSDUH 2024 sample. All categories other than "Hispanic" are specifically non-Hispanic individuals who identify with the given category. 

### Draft 1

```{r}
#| fig-width: 10
#| fig-height: 7
#| out-width: '100%'

ggplot(nsduh2024, aes(x = newrace2)) +
  geom_bar(fill = "steelblue") +
  theme_minimal() +
  labs(title = "Count of Race/Ethnicity", x = "Race/Ethnicity", y = "Count")
```

### Draft 2

```{r}
#| fig-width: 10
#| fig-height: 7
#| out-width: '100%'

#edit new visual, I had to put this as a placeholder for the tab name to work properly

ggplot(nsduh2024, aes(x = newrace2)) +
  geom_bar(fill = "steelblue") +
  theme_minimal() +
  labs(title = "2024 NSDUH Data Subset Race/Ethnicity", x = "Race/Ethnicity", y = "Count") +
  theme(axis.text.x = element_text(
      angle = 45,        
      hjust = 1,        
      vjust = 1))

```

# AUD and Poverty Level

## Column (.panel width="70%")

### Final Version

```{r}
#| echo: false
##variable preparation

aud_pov <- nsduh2024 %>%
  group_by(poverty3, irpyud5alc) %>%
  summarize(aud_count = n()) %>%
  mutate(aud_pct = aud_count / sum(aud_count))
aud_pov <- as.data.frame(aud_pov)

#from okabe ito
custom_colors <- c("Yes" = "#F5C710", "No" = "#56B4E9")
```

```{r}
#| title: 'How did rates of past-year AUD vary by poverty level in 2024?'
#| fig-width: 10
#| fig-height: 7
#| out-width: '100%'

ggplot(na.omit(aud_pov), aes(x = poverty3, y = aud_count, fill = irpyud5alc)) +
  geom_col() +
  geom_label(
    aes(label = paste0(round(aud_pct * 100, 0), "%")),
    vjust = "center",
    hjust = "center",
    size = 4,
    fill = 'white') + 
  ylim(0, 40000) +
  scale_y_continuous(labels = label_comma()) +
  theme_minimal() +
  labs(title = "About 1 in 10 respondents had AUD across poverty level in 2024", x = "Poverty Level", y = "Count",
       fill = "Alcohol Use Disorder", subtitle = "Proportions of past-year AUD rates were nearly equal at each poverty level", caption = 'Source: 2024 NSDUH') +
   scale_fill_manual(values = custom_colors) +
   theme(legend.position = "top",
        legend.key.size = unit(1, 'cm'), #change legend key size
        legend.key.height = unit(1, 'cm'), #change legend key height
        legend.key.width = unit(1, 'cm'), #change legend key width
        legend.title = element_text(size=14), #change legend title font size
        legend.text = element_text(size=10),
        panel.grid.major.x = element_blank(),
        plot.title = element_text(size = 16, face = "bold"),
        plot.subtitle = element_text(size = 16),
        axis.title.x = element_text(size = 14, face = "bold", margin = margin(t = 10, unit = "pt")),
        axis.title.y = element_text(size = 14, face = "bold", margin = margin(r = 10, unit = "pt")),
        axis.text.x = element_text(size = 12, face = "bold"),
        axis.text.y = element_text(size = 12),
        plot.caption = element_text(size = 10, color = "gray50", hjust = 0))


```

## Column {.tabset width="30%"}

### About

One of the demographic pieces of information captured by the NSDUH is poverty status. The data set also provides information on alcohol use disorder. We wanted to see the breakdown of alcohol use disorder by poverty status, which is shown to the left. Notably, the proportions of respondents in each poverty level category who are identified as having AUD are consistent.

### Draft 1

```{r}
#| fig-width: 10
#| fig-height: 7
#| out-width: '100%'

ggplot(na.omit(nsduh2024), aes(x = poverty3, fill = irpyud5alc)) +
  geom_bar() +
  theme_minimal() +
  labs(title = "Alcohol Use Disorder by Poverty Level", x = "Poverty Level", y = "Count",
       fill = "Alcohol Use Disorder")
```

### Draft 2

```{r}
#| fig-width: 10
#| fig-height: 7
#| out-width: '100%'

ggplot(na.omit(nsduh2024), aes(x = poverty3, fill = irpyud5alc)) +
  geom_bar() +
  theme_minimal() +
  labs(title = "Alcohol Use Disorder by Poverty Level", x = "Poverty Level", y = "Count",
       fill = "Alcohol Use Disorder") +
  scale_fill_brewer(palette = "Dark2") +
  theme(legend.position = "none",
        panel.grid.major.x = element_blank(),
        plot.title = element_text(size = 16, face = "bold"),
        axis.title = element_text(size = 12, face = "bold"),
        axis.text.x = element_text(size = 10),
        plot.caption = element_text(size = 9, color = "gray50", hjust = 0))

```

### Draft 3

```{r}
ggplot(na.omit(nsduh2024), aes(irpyud5alc)) +
  geom_bar(aes(fill = poverty3), stat = "count") +
  theme_minimal() +
  labs(title = "Alcohol Use Disorder by Poverty Level", x = "Alcohol Use Disorder", y = "Count",
       fill = "Poverty Level", subtitle = "Across Poverty Levels, Proportions of Respondents with AUD Remains the Same") +
  scale_fill_brewer(palette = "Dark2") +
  theme(legend.position = c(.8, .8),
        legend.key.size = unit(1, 'cm'), #change legend key size
        legend.key.height = unit(1, 'cm'), #change legend key height
        legend.key.width = unit(1, 'cm'), #change legend key width
        legend.title = element_text(size=14), #change legend title font size
        legend.text = element_text(size=10),
        panel.grid.major.x = element_blank(), 
        plot.title = element_text(size = 16, face = "bold"),
        axis.title = element_text(size = 14, face = "bold"),
        axis.text.x = element_text(size = 12), 
        axis.text.y = element_text(size = 12),
        plot.caption = element_text(size = 9, color = "gray50", hjust = 0))
```


# AUD by Age Group

```{r}
#kat
#setting up

# get proportion of age group with aud
aud_pct_age <- nsduh2024 |>
  group_by(catag6, irpyud5alc) |>
  summarize(aud_count = n()) |>
  mutate(aud_pct = aud_count / sum(aud_count)) |>
  filter(irpyud5alc == "Yes")

aud_pct_age_f <- aud_pct_age

# shorten catag6 values for final viz
aud_pct_age_f$catag6 <- recode(
  aud_pct_age_f$catag6,
    '12-17 Years Old' = '12 to 17', 
    '18-25 Years Old' = '18 to 25', 
    '26-34 Years Old' = '26 to 34', 
    '34-49 Years Old' = '34 to 49',
    '50-64 Years Old' = '50 to 64',
    '65 or Older' = '65+')

```

## Column (.panel width="70%")

### Final Version

```{r}
#| title: How did rates of past-year AUD prevalence vary across age groups in 2024?
#| fig-width: 10
#| fig-height: 7
#| out-width: '100%'

aud_pct_age_f |>
  ggplot(aes(catag6, aud_pct, fill = catag6)) +
  geom_col(position = 'dodge') +
  geom_text_repel(aes(label = scales::percent(aud_pct, accuracy = 0.1)),
                   nudge_y = 0.001,
                   segment.color = NA,
                   direction = 'y') +
  scale_y_continuous(labels = scales::percent, limits = c(0, .15)) +
  #coord_flip() +
  labs(title = 'AUD Rates are Highest Among Young Adult Age Groups in 2024',
       subtitle = "Approximately 3 out of 10 young adults ages <span style='color:#D95F02;'>**18 to 25**</span> and <span style='color:#7570B3;'>**26 to 34**</span> had past-year AUD combined",
       caption = 'Source: 2024 NSDUH',
       x = 'Age (Years)',
       y = 'Past-Year AUD (Percentage)') +
  theme_minimal() +
  scale_fill_brewer(palette = "Dark2") +
  theme(legend.position = "none",
        panel.grid.major.x = element_blank(),
        plot.title = element_text(size = 18, face = "bold"),
        plot.subtitle = element_markdown(), element_text(size = 16),
        axis.title.x = element_text(size = 14, face = "bold", margin = margin(t = 10, unit = "pt")),
        axis.text.x = element_text(size = 12, face = "bold"),
        axis.title.y = element_text(size = 14, face = "bold", margin = margin(r = 10, unit = "pt")),
        axis.text.y = element_text(size = 12, face = "bold"),
        plot.caption = element_text(size = 10, color = "gray50", hjust = 0)
       )
```

## Column {.tabset width="30%"}

### About

This figure shows the percentage of participants that were classified as having alcohol use disorder in the past year within each age group.

First, the data was grouped by age group and AUD status (i.e., No or Yes). The count for each Age Group by AUD status group and the percentage of participants within each age group by AUD status were calculated. Finally, the data was filtered to only include people with AUD (*n* = 6,069) to examine the distribution of AUD prevalence across age groups.

Based on the visualization, the highest prevalence rates of AUD were in the 26 to 34 and 18 to 25 year old age groups, at 14.1% and 14.0%, respectively. This pattern aligns with the literature, which suggests that AUD peaks in young adulthood, then declines across time.

### Draft 1

```{r}
#| title: Number of People with Past-Year Alcohol Use Disorder by Age Group in 2024
#| fig-width: 10
#| fig-height: 7
#| out-width: '100%'

nsduh_sud |>
  group_by(catag6) |>
  ggplot(aes(catag6, aud_count)) +
  geom_col() +
  labs(title = 'Number of People with Past-Year Alcohol Use Disorder by Age Group in 2024',
       x = 'Age Group',
       y = 'Count') +
  theme_minimal()

```

### Draft 2

```{r}
#| title: Proportion of People with Past-Year Alcohol Use Disorder in 2024 by Age Group
#| fig-width: 10
#| fig-height: 7
#| out-width: '100%'

aud_pct_age |>
  ggplot(aes(catag6, aud_pct)) +
  geom_col(position = 'dodge') +
  scale_y_continuous(labels = scales::percent) +
  coord_flip() +
  labs(title = 'Proportion of People with Past-Year Alcohol Use Disorder in 2024 by Age Group',
       x = 'Age Group',
       y = 'Percentage with AUD') +
  theme_minimal()

```

# Alcohol Use Treatment

```{r}
# kat
# setting up

# turn classified as needing substance use treatment for alc use - past year into a factor

nsduh2024$sutnedalcpy2 <- factor(
  nsduh2024$sutnedalcpy2,
  levels = c('0', '1'),
  labels = c('No', 'Yes'))

# turn received substance use treatment for alc use - past year into a factor

nsduh2024$sutrtalcpy2 <- factor(
  nsduh2024$sutrtalcpy2,
  levels = c('0', '1'),
  labels = c('No', 'Yes'))

#dataset with people who needed su treatment for alc use in the past year
nsduh_need_alc <- nsduh2024 |>
  filter(sutnedalcpy2 == 'Yes') |>
  group_by(sutrtalcpy2) |>
  summarize(treat_count = n()) |>
  mutate(treat_pct = treat_count / sum(treat_count))

```

## Column (.panel width="70%")

### Final Version

```{r}
#| title: 'How many people who needed alcohol use treatment in 2024 actually received some form of treatment?'
#| fig-width: 10
#| fig-height: 7
#| out-width: '100%'

nsduh_need_alc |>
  ggplot(aes(sutrtalcpy2, treat_pct, fill = sutrtalcpy2)) +
  geom_col() +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  geom_text_repel(aes(label = scales::percent(treat_pct, accuracy = 0.1)),
                   nudge_y = 0.001,
                   segment.color = NA,
                   direction = 'y') +
  labs(title = 'Most Alcohol Use Treatment Needs Were Unmet in 2024',
       subtitle = 'Only about 1 in 10 people with AUD received any form of treatment',
       x = 'Received Treatment',
       y = 'Needed Treatment (Percentage)',
       caption = 'Source: 2024 NSDUH') +
  theme_minimal() +
  scale_fill_brewer(palette = "Set1") +
  theme(legend.position = "none",
        panel.grid.major.x = element_blank(),
        plot.title = element_text(size = 18, face = "bold"),
        plot.subtitle = element_markdown(), element_text(size = 16),
        axis.title.x = element_text(size = 14, face = "bold", margin = margin(t = 10, unit = "pt")),
        axis.text.x = element_text(size = 12, face = "bold"),
        axis.title.y = element_text(size = 14, face = "bold", margin = margin(r = 10, unit = "pt")),
        axis.text.y = element_text(size = 12, face = "bold"),
        plot.caption = element_text(size = 10, color = "gray50", hjust = 0)
       )

```

## Column {.tabset width="30%"}

### About

This figure shows the percentage of people who received substance use treatment for their alcohol use (*n* = 934) among participants who were classified as needing treatment (i.e., having alcohol use disorder; *n* = 6,069).

First, the data was filtered to only include participants who were classified as needing alcohol use treatment due to having AUD in the past year. Then, the data was grouped by whether treatment was received (i.e., No or Yes). The percentage of participants who needed treatment by treatment status group was calculated. 

Based on the visualization, the vast majority of participants who needed alcohol use treatment did not receive any form of treatment in 2024. It is important to note that this percentage does not take into account "successful" treatment, or whether treatment was continued beyond one session or resource accessed. 

### Draft 1

```{r}
#| title: 'Past-Year Alcohol Use Treatment Needs Met in 2024'
#| fig-width: 10
#| fig-height: 7
#| out-width: '100%'

nsduh_need_alc |>
  ggplot(aes(sutrtalcpy2, treat_pct)) +
  geom_col() +
  scale_y_continuous(labels = scales::percent) +
  labs(title = 'Past-Year Alcohol Use Treatment Needs Met in 2024',
       subtitle = 'Among individuals who were classified as needing treatment for their alcohol use in the 2024 NSDUH',
       x = 'Received Treatment',
       y = 'Treatment Needed (Percentage)') +
  theme_minimal()

```

### Draft 2

```{r}
#| fig-width: 10
#| fig-height: 7
#| out-width: '100%'

nsduh_need_alc |>
  ggplot(aes(sutrtalcpy2, treat_pct, fill = sutrtalcpy2)) +
  geom_col() +
  scale_y_continuous(labels = scales::percent) +
  geom_label_repel(aes(label = scales::percent(treat_pct, accuracy = 0.1)),
                   nudge_y = 0.05,
                   fill = NA) +
  labs(title = 'Past-Year Alcohol Use Treatment Needs Met in 2024',
       x = 'Received Treatment',
       y = 'Treatment Needed (Percentage)') +
  theme_minimal() +
  theme(legend.position = "none")

```
